---
title: "Hands-on_Ex04a"
format: html
editor: visual
---

# **1 Learning Outcome**

Visualising distribution is not new in statistical analysis. In chapter 1 we have shared with you some of the popular statistical graphics methods for visualising distribution are histogram, probability density curve (pdf), boxplot, notch plot and violin plot and how they can be created by using ggplot2. In this chapter, we are going to share with you two relatively new statistical graphic methods for visualising distribution, namely ridgeline plot and raincloud plot by using ggplot2 and its extensions.

# **2 Getting Started**

::: panel-tabset
## Installing and loading the packages

```{r}
pacman::p_load(ggdist, ggridges, ggthemes,
               colorspace, tidyverse)
```

## Data import

```{r}
exam <- read.csv("data/Exam_data.csv")
```
:::

# **3 Visualising Distribution with Ridgeline Plot**

Ridgeline plot (sometimes called Joyplot) is a data visualisation technique for revealing the distribution of a numeric value for several groups. Distribution can be represented using histograms or density plots, all aligned to the same horizontal scale and presented with a slight overlap.

::: callout-note
-   Ridgeline plots make sense when the number of group to represent is medium to high, and thus a classic window separation would take to much space. Indeed, the fact that groups overlap each other allows to use space more efficiently. If you have less than 5 groups, dealing with other distribution plots is probably better.

-   It works well when there is a clear pattern in the result, like if there is an obvious ranking in groups. Otherwise group will tend to overlap each other, leading to a messy plot not providing any insight.
:::

## **3.1 Plotting ridgeline graph: ggridges method**

There are several ways to plot ridgeline plot with R. In this section, you will learn how to plot ridgeline plot by using [ggridges](https://wilkelab.org/ggridges/index.html) package.

ggridges package provides two main geom to plot gridgeline plots, they are: [`geom_ridgeline()`](https://wilkelab.org/ggridges/reference/geom_ridgeline.html) and [`geom_density_ridges()`](https://wilkelab.org/ggridges/reference/geom_density_ridges.html). The former takes height values directly to draw the ridgelines, and the latter first estimates data densities and then draws those using ridgelines.

The ridgeline plot below is plotted by using `geom_density_ridges()`.

### **3.1.1 `geom_density_ridges`**

::: panel-tabset
## The plot

```{r,echo=FALSE}
ggplot(exam, 
       aes(x = ENGLISH, 
           y = CLASS)) +
  geom_density_ridges(
    scale = 3,
    rel_min_height = 0.01,
    bandwidth = 3.4,
    fill = lighten("#7097BB", .3),
    color = "white"
  ) +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
    ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```

## The code Chunk

```{r,eval=FALSE}
ggplot(exam, 
       aes(x = ENGLISH, 
           y = CLASS)) +
  geom_density_ridges(
    scale = 3,
    rel_min_height = 0.01,
    bandwidth = 3.4,
    fill = lighten("#7097BB", .3),
    color = "white"
  ) +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
    ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```
:::

::: callout-note
### **Insights from the Ridgeline Plot**

1.  **Grade Distribution Across Classes**

    -   The graph shows the distribution of **English grades** across different class sections (e.g., 3A, 3B, 3C, etc.).

    -   Each ridgeline represents a different class, allowing comparison of grade distributions.

2.  **Grade Trends and Variability**

    -   Most classes have a **peak around 50-75** on the grade scale, indicating that most students score in this range.

    -   Some distributions have **wider spreads**, suggesting higher variability in student performance.
:::

## **3.2 Varying fill colors along the x axis**

Sometimes we would like to have the area under a ridgeline not filled with a single solid color but rather with colors that vary in some form along the x axis. This effect can be achieved by using either [`geom_ridgeline_gradient()`](https://wilkelab.org/ggridges/reference/geom_ridgeline_gradient.html) or [`geom_density_ridges_gradient()`](https://wilkelab.org/ggridges/reference/geom_ridgeline_gradient.html). Both geoms work just like `geom_ridgeline()` and `geom_density_ridges()`, except that they allow for varying fill colors. However, they do not allow for alpha transparency in the fill. For technical reasons, we can have changing fill colors or transparency but not both.

::: panel-tabset
## The plot

```{r,echo=FALSE}
ggplot(exam, 
       aes(x = ENGLISH, 
           y = CLASS,
           fill = stat(x))) +
  geom_density_ridges_gradient(
    scale = 3,
    rel_min_height = 0.01) +
  scale_fill_viridis_c(name = "Temp. [F]",
                       option = "C") +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
  ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```

## The code chunk

```{r,eval=FALSE}
ggplot(exam, 
       aes(x = ENGLISH, 
           y = CLASS,
           fill = stat(x))) +
  geom_density_ridges_gradient(
    scale = 3,
    rel_min_height = 0.01) +
  scale_fill_viridis_c(name = "Temp. [F]",
                       option = "C") +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
  ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```
:::

:::: callout-note
::: callout-tip
**Insights**:

This ridgeline plot visualizes the distribution of English scores across different classes, using a light-to-dark color gradient to represent increasing scores. Most distributions peak between 50 and 80, indicating a common performance range. The overlapping design makes it easy to compare trends across groups, though some areas may be difficult to distinguish due to excessive overlap. A shared legend for color intensity and slight adjustments to opacity or spacing could further enhance readability.
:::
::::

### **3.3 Mapping the probabilities directly onto colour**

Beside providing additional geom objects to support the need to plot ridgeline plot, ggridges package also provides a stat function called [`stat_density_ridges()`](https://wilkelab.org/ggridges/reference/stat_density_ridges.html) that replaces [`stat_density()`](https://ggplot2.tidyverse.org/reference/geom_density.html) of ggplot2.

Figure below is plotted by mapping the probabilities calculated by using `stat(ecdf)` which represent the empirical cumulative density function for the distribution of English score.

::: panel-tabset
## The plot

```{r,echo=FALSE}
ggplot(exam,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = 0.5 - abs(0.5-stat(ecdf)))) +
  stat_density_ridges(geom = "density_ridges_gradient", 
                      calc_ecdf = TRUE) +
  scale_fill_viridis_c(name = "Tail probability",
                       direction = -1) +
  theme_ridges()+
  labs(x = "English",
       y = "",
       title="Uneven distribution across classes.") +   
  theme(
    #legend.position = "none",
    #panel.grid.major = element_blank(),
    axis.title.y = element_blank(),
    plot.background = element_rect(fill="#f5f5f5",colour="#f5f5f5")  
    ) 
```

## The code chunk

```{r,eval=FALSE}
ggplot(exam,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = 0.5 - abs(0.5-stat(ecdf)))) +
  stat_density_ridges(geom = "density_ridges_gradient", 
                      calc_ecdf = TRUE) +
  scale_fill_viridis_c(name = "Tail probability",
                       direction = -1) +
  theme_ridges()+
  labs(x = "English",
       y = "",
       title="Uneven distribution across classes.") +   
  theme(
    #legend.position = "none",
    #panel.grid.major = element_blank(),
    axis.title.y = element_blank(),
    plot.background = element_rect(fill="#f5f5f5",colour="#f5f5f5")  
    ) 
```
:::

::: callout-tip
**Insights:**

-   The distribution of English scores varies across classes, with some showing a more even spread while others have concentrated peaks. The color gradient represents tail probabilities, highlighting areas of lower density and potential outliers.
:::

### **3.4 Ridgeline plots with quantile lines**

::: panel-tabset
## The plot

```{r,echo=FALSE}
ggplot(exam,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = factor(stat(quantile))
           )) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE, 
    quantiles = 4,
    quantile_lines = TRUE) +
  scale_fill_viridis_d(name = "Quartiles") +
  theme_ridges()
```

## The code chunk

```{r,eval=FALSE}
ggplot(exam,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = factor(stat(quantile))
           )) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE, 
    quantiles = 4,
    quantile_lines = TRUE) +
  scale_fill_viridis_d(name = "Quartiles") +
  theme_ridges()
```
:::

Instead of using number to define the quantiles, we can also specify quantiles by cut points such as 2.5% and 97.5% tails to colour the ridgeline plot as shown in the figure below.

::: panel-tabset
## The plot

```{r,echo=FALSE}
ggplot(exam,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = factor(stat(quantile))
           )) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE, 
    quantiles = c(0.025, 0.975)
    ) +
  scale_fill_manual(
    name = "Probability",
    values = c("#FF0000A0", "#A0A0A0A0", "#0000FFA0"),
    labels = c("(0, 0.025]", "(0.025, 0.975]", "(0.975, 1]")
  ) +
  theme_ridges()
```

## The code chunk

```{r,eval=FALSE}
ggplot(exam,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = factor(stat(quantile))
           )) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE, 
    quantiles = c(0.025, 0.975)
    ) +
  scale_fill_manual(
    name = "Probability",
    values = c("#FF0000A0", "#A0A0A0A0", "#0000FFA0"),
    labels = c("(0, 0.025]", "(0.025, 0.975]", "(0.975, 1]")
  ) +
  theme_ridges()
```
:::

# **4 Visualizing Distribution with Raincloud Plot**

Raincloud Plot is a data visualisation techniques that produces a half-density to a distribution plot. It gets the name because the density plot is in the shape of a “raincloud”. The raincloud (half-density) plot enhances the traditional box-plot by highlighting multiple modalities (an indicator that groups may exist). The boxplot does not show where densities are clustered, but the raincloud plot does!

In this section, you will learn how to create a raincloud plot to visualise the distribution of English score by race. It will be created by using functions provided by **ggdist** and ggplot2 packages.

## **4.1 Plotting a Half Eye graph**

First, we will plot a Half-Eye graph by using [`stat_halfeye()`](https://mjskay.github.io/ggdist/reference/stat_halfeye.html) of **ggdist** package.

This produces a Half Eye visualization, which is contains a half-density and a slab-interval.

We remove the slab interval by setting .width = 0 and point_colour = NA.

::: panel-tabset
## The plot

```{r,echo=FALSE}

ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA)
```

## The code chunk

```{r,eval=FALSE}
ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA)
```
:::

## **4.2 Adding the boxplot with `geom_boxplot()`**

Next, we will add the second geometry layer using [`geom_boxplot()`](https://r4va.netlify.app/chap09) of ggplot2. This produces a narrow boxplot. We reduce the width and adjust the opacity.

::: panel-tabset
## The plot

```{r,echo=FALSE}

ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA)
```

## The code chunk

```{r,eval=FALSE}

ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA)
```
:::

## **4.3 Adding the Dot Plots with `stat_dots()`**

Next, we will add the third geometry layer using [`stat_dots()`](https://mjskay.github.io/ggdist/reference/stat_dots.html) of ggdist package. This produces a half-dotplot, which is similar to a histogram that indicates the number of samples (number of dots) in each bin. We select side = “left” to indicate we want it on the left-hand side.

::: panel-tabset
## The plot

```{r,echo=FALSE}

ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA) +
  stat_dots(side = "left", 
            justification = 1.2, 
            binwidth = .5,
            dotsize = 2)
```

## The code chunk

```{r,eval=FALSE}

ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA) +
  stat_dots(side = "left", 
            justification = 1.2, 
            binwidth = .5,
            dotsize = 2)
```
:::

### **4.4 Finishing touch**

Lastly, [`coord_flip()`](https://ggplot2.tidyverse.org/reference/coord_flip.html) of ggplot2 package will be used to flip the raincloud chart horizontally to give it the raincloud appearance.

::: panel-tabset
## The plot

```{r,echo=FALSE}

ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA) +
  stat_dots(side = "left", 
            justification = 1.2, 
            binwidth = .5,
            dotsize = 1.5) +
  coord_flip() +
  theme_economist()
```

## The code chunk

```{r,eval=FALSE}

ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA) +
  stat_dots(side = "left", 
            justification = 1.2, 
            binwidth = .5,
            dotsize = 1.5) +
  coord_flip() +
  theme_economist()
```
:::

# **Reference**

-   [Introducing Ridgeline Plots (formerly Joyplots)](https://blog.revolutionanalytics.com/2017/07/joyplots.html)

-   Claus O. Wilke [Fundamentals of Data Visualization](https://clauswilke.com/dataviz/) especially Chapter [6](https://clauswilke.com/dataviz/visualizing-amounts.html), [7](https://clauswilke.com/dataviz/histograms-density-plots.html), [8](https://clauswilke.com/dataviz/ecdf-qq.html), [9](https://clauswilke.com/dataviz/boxplots-violins.html) and [10](https://clauswilke.com/dataviz/visualizing-proportions.html).

-   Allen M, Poggiali D, Whitaker K et al. [“Raincloud plots: a multi-platform tool for robust data. visualization”](https://wellcomeopenresearch.org/articles/4-63) \[version 2; peer review: 2 approved\]. Welcome Open Res 2021, pp. 4:63.

-   [Dots + interval stats and geoms](https://mjskay.github.io/ggdist/articles/dotsinterval.html)
